Data exploration through dot-driven development\footnote{This work was supported by Google Digital News Initiative.}

### 2017-01-12 pivot/paper.tex
References and related work

\EventEditors{John Q. Open and Joan R. Acces}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}

@@

Spreadsheets made data exploration accessible to a large number of people, but operations 
performed on spreadsheets cannot be reproduced or replicated with different input parameters.
The manual mode of interaction is not repeatable and it breaks the link with the original data 
source, making spreadsheets error-prone [X]. The answer to this problem is to explore data 
programmatically. A program can be run repeatedly and its parameters can be modified.

@@

\item Our mechanism illustrates how to embed fancy types [X] into a simple nominally-typed programming  
  language (Section~\ref{sec:columns}). We track names and types of available columns of the 
  manipulated data set (using a mechanism akin to row types), but the same mechanism can be used for 
  embedding other advanced typing schemes into any Java-like language.

@@

\subparagraph{Information-rich programming.} Type providers were first presented as a mechanism
for providing type-safe access to rich information sources. A type provider is a compile-time 
component that imports external information source into a programming language [X]. It provides two 
things to the compiler or editor hosting it: a type signature that models the external source 
using structures understood by the host language (e.g.~types) and an implementation for the 
signatures which accesses data from the external source.

For example, the World Bank type provider [X] provides a fine-grained access to development 
indicators about countries. The following accesses CO2 emissions by~country~in~2010:
%
\begin{equation*}
\begin{array}{l}
\ident{world}.\ident{byYear}.\qident{2010}.\qident{Climate Change}.\qident{CO2 emissions (kt)}
\end{array}
\end{equation*}

@@

\subparagraph{Types from data.} Recent work on the F\# Data library [X] uses type providers for 
accessing data in structured formats such as XML, CSV and JSON. This is done by inferring the 
structure of the data from a sample document, provided as a static parameter to a type provider.
In the following example, adapted from [X], a sample URL is passed to \ident{JsonProvider}:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Weather} = \ident{JsonProvider}\langl\str{http://api.owm.org/?q=London}\rangl \\[0.5em]
 \kvd{let}~\ident{ldn} = \ident{Weather.GetSample}()\\
 \ident{printfn}~\str{The temperature in London is \%f}~\ident{ldn.Main.Temp}
\end{array}
\end{equation*}

@@

Type providers in F\# rely on dot-driven development when navigating through data. When writing code 
to access current temperature $\ident{ldn.Main.Temp}$ in Section~\ref{sec:tps}, the auto-completion
offers various available properties, such as \ident{Wind} and \ident{Clouds} once ``dot'' is typed
after $\ident{ldn.Main}$. Other type providers [X] follow a similar pattern. It is worth noting
that despite the use of nominal typing, the names of types rarely explicitly appear in code -- we
do not need to know the name of the type of $\ident{ldn.Main}$, but we need to know its members.
Thus the type name can be arbitrary [X] and is used more as a lookup key.

@@

\subparagraph{Values vs. types.}
As the Figure~\ref{fig:dot-driven} illustrates, the pivot type provider sometimes blurs the 
distinction between values and (members of) types. In the example in Section~\ref{sec:intro},
\str{Rio (2016)} is a string value in Python, but a statically-typed member \qident{Rio (2016)}
when using pivot type provider. This is a recurring theme in type provider development\footnote{The
\ident{Individuals} property in the Freebase type provider [X] imports values into types in a similar way.}.

@@

Unlike in traditional development, a data scientist doing data exploration often has the entire 
data set available. The pivot type provider uses this when offering possible values for filtering
(Section~X), but all other operations (Section~Y) require only meta-data (names and types of 
columns). Following the example of type providers for structured data formats [X], the schema could
be inferred from a representative sample.

@@

% https://github.com/fsharp/fslang-design/blob/master/FSharp-4.0/StaticMethodArgumentsDesignAndSpec.md

In F\# type providers, this could be done using methods with static parameters [X] by writing, 
for example, $\ident{g}.\qident{sum Gold as}\langl\str{Total Gold}\rangl()$. In F\#, the value of 
the static parameter (here, \str{Total Gold}) is passed to the type provider, which can use it to
generate the type signature of the method and the return type with member name according to the
value of the static parameter.

@@

\noindent
Operations on the type of rows in the data set can be captured using row types [X]. In addition, 
we need to annotate type with a form of type state [Y] to restrict what operations are available. 
When dropping columns, we first access the \qident{drop columns} member, which sets the state to
a state where we can drop individual columns using \qident{drop $f$}. The \ident{then} member can 
then be used to complete the operation and choose another transformation.

@@

\noindent
The rule, adapted from [X], does not capture laziness of type providers that is 
important for the pivot type provider (where the number of provided classes is potentially 
infinite). We discuss this aspect in Section~\ref{sec:foo}.

@@

We do not fully develop the type system based on fancy types sketched in Section~\ref{sec:columns-row}.
However, the remark illustrates one interesting aspect of our work -- the type provider 
mechanism makes it possible to express safety guarantees that would normally require row types and 
type state in a simple nominally typed language. In a similar way, type providers have been used
to encode session types [X], suggesting that this is a generally useful approach.

@@

The focus of our work is on data aggregation and so we use a form of relational algebra with 
extensions for grouping and sorting [X,Y,Z]. The syntax is defined in Figure~\ref{fig:foo-rel}.
We write $f$ for column (field) names and we include definition of a data value $D$, which maps
columns to vectors of length $r$ storing the data (values $v$ are defined below).
Aside from standard projection $\Pi$ and selection $\sigma$, our algebra includes sorting 
$\tau$ which takes one or more columns forming the sort key (with sort order $\omega$)
and aggregation $\Phi$, which requires a single grouping key and several aggregations together with 
names of the new columns to be returned.

@@

We model the host language using a variant of the Foo calculus [X]. The core of the calculus models
a simple object-based language with objects and members. The syntax of the language is shown in
Figure~\ref{fig:foo-syntax}. The relational algebra defined in Figure~\ref{fig:foo-rel} is included
in the Foo calculus as a model of the runtime components of the pivot type provider -- the values
include the data value $D$ and the expressions include all the operations of the relational algebra.

@@

\subparagraph{Type checking.} One interesting aspect of type checking with type providers is that
type providers can provide potentially infinite number of types. In F\#, the types are provided 
lazily as the type checker explores parts of the type space used by the program [X]. Consider:
%
\begin{equation*}
\ident{olympics}.\qident{group data}.\qident{by Athlete}.\qident{sum Gold}.\ident{then}
\end{equation*}

\noindent
The type checker initially knows the type of $\ident{olympics}$ is a class $C_1$ with member 
$\qident{group data}$ and it knows that the type of this member is $C_2$. However, it only needs to 
obtain full definition of $C_2$ when checking the member $\qident{by Athlete}$. Types of other 
member of $C_1$ remain unevaluated when they do not appear in the source code. This aspect of type
providers have been omitted in previous work [X,Y], but it is necessary for the pivot type provider.
The typing rules given are written using the following judgement:
%
\begin{equation*}
L_1; \Gamma \vdash e : \tau; L_2
\end{equation*}

@@

\noindent
The class $C$ may use classes defined in $L_1$, but also delayed classes from $L$. This models 
laziness as $L$ is a function that may never be evaluated. Since $L$ is potentially infinite, we 
cannot check class definitions upfront as in typical object calculi [X]. Instead, we check that 
that members are well typed as they appear in the source code, which matches the behaviour of F\# 
type providers. In general, this means that $L$ may contain classes with incorrectly typed 
members. We prove that this is not the case for the pivot type provider (Section~\ref{sec:pivot-prop}).

@@

The rules for primitive types and variables are standard. Input data (\emph{data}) is of type 
\ident{Query} and all the operations of relational algebra take a value of type \ident{Query} and 
produce a result of type \ident{Query}. An untyped \ident{Query} value can be converted into a 
series type (\emph{series}) of any type -- in the pivot type provider, the \ident{Query} type is 
always hidden and users can only work with type-safe series type. The bridging of typed and 
untyped parts of the language resembles the boundary between static and dynamic typing in 
gradually typed languages.% [X,Y,Z].

@@

In an implementation, a type provider is an executable component that is called by the compiler
and the editor to provide information about types on demand. In our formalization, we follow the 
style used by Petricek et al. [X], but we add laziness as discussed in Section~\ref{sec:foo-foo}.
We model the core operations (dropping columns, grouping and sorting) in Section~\ref{sec:pivot-core}
and then refine the model to include filtering based on values Section~\ref{sec:pivot-filter}.
We omit paging, as it does not affect the shape of data, but modelling it would require richer 
host language.

@@

\noindent
In prior work on type providers [X], the actual input value is not available when writing the code
and so the schema is inferred from a representative sample value. In exploratory data analysis, the 
data set is often available at the time of writing the code and so $D$ can be the actual data set.
However, it could equally be a sample or representative subset of the data.

@@

\subparagraph{Spreadsheet-inspired live editor.}
The third editor extension that we developed for the pivot type provider aims to bridge the gap 
between code and user interface. This is done through a direct manipulation interface [X] inspired
by spreadsheet applications. When exploring data in a spreadsheet, the user can always see the 
data they work with and the results of an action will be immediately visible. For queries written
using the pivot type provider, the results can also be calculated immediately using the sample 
data set provided when using the refined version of the type provider as discussed in 
Section~\ref{sec:pivot-filter}.

@@

Unlike the user interface for specifying lists, the live editor has been implemented specifically
for the pivot type provider, but it still heavily relies on the simple structure provided by the 
fact that entire transformation can be written as a single chain of member accesses. In particular,
we identify individual transformations (\qident{group by}, \qident{sort by}, etc.) and then 
generate different user interface for specifying parameters. For sorting, as shown in 
Figure~\ref{fig:case-ed}, the second level lets the user add or remove members. For grouping or
paging, the second level looks different and lets the user choose the key and the number of 
elements to take, respectively.

@@

a

b

c

\subsection{Further work}
\label{sec:further}

\subsection{Related work}

visual query systems

https://people.csail.mit.edu/ebakke/research/sieuferd_sigmod2016.pdf

\subsection{Conclusions}

\subparagraph*{Acknowledgements.}

I want to thank \dots

@@

\bibliography{lipics-v2016-sample-article}

>>>>

\EventEditors{John Q. Open and Joan R. Acces}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{DRAFT}
\EventAcronym{DRAFT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}

@@

Spreadsheets made data exploration accessible to a large number of people, but operations 
performed on spreadsheets cannot be reproduced or replicated with different input parameters.
The manual mode of interaction is not repeatable and it breaks the link with the original data 
source, making spreadsheets error-prone \cite{exceldep}. The answer to this problem is to explore data 
programmatically. A program can be run repeatedly and its parameters can be modified.

@@

\item Our mechanism illustrates how to embed fancy types \cite{fancytypes} into a simple nominally-typed programming  
  language (Section~\ref{sec:columns}). We track names and types of available columns of the 
  manipulated data set (using a mechanism akin to row types), but the same mechanism can be used for 
  embedding other advanced typing schemes into any Java-like language.

@@

\subparagraph{Information-rich programming.} Type providers were first presented as a mechanism
for providing type-safe access to rich information sources. A type provider is a compile-time 
component that imports external information source into a programming language \cite{inforich}. It provides two 
things to the compiler or editor hosting it: a type signature that models the external source 
using structures understood by the host language (e.g.~types) and an implementation for the 
signatures which accesses data from the external source.

For example, the World Bank type provider \cite{ageweb} provides a fine-grained access to development 
indicators about countries. The following accesses CO2 emissions by~country~in~2010:
%
\begin{equation*}
\begin{array}{l}
\ident{world}.\ident{byYear}.\qident{2010}.\qident{Climate Change}.\qident{CO2 emissions (kt)}
\end{array}
\end{equation*}

@@

\subparagraph{Types from data.} Recent work on the F\# Data library \cite{fsdata} uses type providers for 
accessing data in structured formats such as XML, CSV and JSON. This is done by inferring the 
structure of the data from a sample document, provided as a static parameter to a type provider.
In the following example, adapted from \cite{fsdata}, a sample URL is passed to \ident{JsonProvider}:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Weather} = \ident{JsonProvider}\langl\str{http://api.owm.org/?q=London}\rangl \\[0.5em]
 \kvd{let}~\ident{ldn} = \ident{Weather.GetSample}()\\
 \ident{printfn}~\str{The temperature in London is \%f}~\ident{ldn.Main.Temp}
\end{array}
\end{equation*}

@@

Type providers in F\# rely on dot-driven development when navigating through data. When writing code 
to access current temperature $\ident{ldn.Main.Temp}$ in Section~\ref{sec:tps}, the auto-completion
offers various available properties, such as \ident{Wind} and \ident{Clouds} once ``dot'' is typed
after $\ident{ldn.Main}$. Other type providers \cite{inforich} follow a similar pattern. It is worth noting
that despite the use of nominal typing, the names of types rarely explicitly appear in code -- we
do not need to know the name of the type of $\ident{ldn.Main}$, but we need to know its members.
Thus the type name can be arbitrary \cite{fsdata} and is used more as a lookup key.

@@

\subparagraph{Values vs. types.}
As the Figure~\ref{fig:dot-driven} illustrates, the pivot type provider sometimes blurs the 
distinction between values and (members of) types. In the example in Section~\ref{sec:intro},
\str{Rio (2016)} is a string value in Python, but a statically-typed member \qident{Rio (2016)}
when using pivot type provider. This is a recurring theme in type provider development\footnote{The
\ident{Individuals} property in the Freebase type provider \cite{inforich} imports values into types in a similar way.}.

@@

Unlike in traditional development, a data scientist doing data exploration often has the entire 
data set available. The pivot type provider uses this when offering possible values for filtering
(Section~X), but all other operations (Section~Y) require only meta-data (names and types of 
columns). Following the example of type providers for structured data formats \cite{fsdata}, the schema could
be inferred from a representative sample.

@@

In F\# type providers, this could be done using methods with static parameters \cite{staticpar} by writing, 
for example, $\ident{g}.\qident{sum Gold as}\langl\str{Total Gold}\rangl()$. In F\#, the value of 
the static parameter (here, \str{Total Gold}) is passed to the type provider, which can use it to
generate the type signature of the method and the return type with member name according to the
value of the static parameter.

@@

\noindent
Operations on the type of rows in the data set can be captured using row types \cite{rowtypes}. 
In addition, we need to annotate type with a form of type state \cite{typestate} to restrict what operations are available. 
When dropping columns, we first access the \qident{drop columns} member, which sets the state to
a state where we can drop individual columns using \qident{drop $f$}. The \ident{then} member can 
then be used to complete the operation and choose another transformation.

@@

\noindent
The rule, adapted from \cite{fsdata}, does not capture laziness of type providers that is 
important for the pivot type provider (where the number of provided classes is potentially 
infinite). We discuss this aspect in Section~\ref{sec:foo}.

@@

We do not fully develop the type system based on fancy types sketched in Section~\ref{sec:columns-row}.
However, the remark illustrates one interesting aspect of our work -- the type provider 
mechanism makes it possible to express safety guarantees that would normally require row types and 
type state in a simple nominally typed language. In a similar way, type providers have been used
to encode session types \cite{sessiontp}, suggesting that this is a generally useful approach.

@@

The focus of our work is on data aggregation and so we use a form of relational algebra with 
extensions for grouping and sorting \cite{relalg,dbsys}. The syntax is defined in Figure~\ref{fig:foo-rel}.
We write $f$ for column (field) names and we include definition of a data value $D$, which maps
columns to vectors of length $r$ storing the data (values $v$ are defined below).
Aside from standard projection $\Pi$ and selection $\sigma$, our algebra includes sorting 
$\tau$ which takes one or more columns forming the sort key (with sort order $\omega$)
and aggregation $\Phi$, which requires a single grouping key and several aggregations together with 
names of the new columns to be returned.

@@

We model the host language using a variant of the Foo calculus \cite{fsdata}. The core of the calculus models
a simple object-based language with objects and members. The syntax of the language is shown in
Figure~\ref{fig:foo-syntax}. The relational algebra defined in Figure~\ref{fig:foo-rel} is included
in the Foo calculus as a model of the runtime components of the pivot type provider -- the values
include the data value $D$ and the expressions include all the operations of the relational algebra.

@@

\subparagraph{Type checking.} One interesting aspect of type checking with type providers is that
type providers can provide potentially infinite number of types. In F\#, the types are provided 
lazily as the type checker explores parts of the type space used by the program \cite{inforich}. 
Consider:
%
\begin{equation*}
\ident{olympics}.\qident{group data}.\qident{by Athlete}.\qident{sum Gold}.\ident{then}
\end{equation*}

\noindent
The type checker initially knows the type of $\ident{olympics}$ is a class $C_1$ with member 
$\qident{group data}$ and it knows that the type of this member is $C_2$. However, it only needs to 
obtain full definition of $C_2$ when checking the member $\qident{by Athlete}$. Types of other 
member of $C_1$ remain unevaluated when they do not appear in the source code. This aspect of type
providers have been omitted in previous work \cite{fsdata,liteq}, but it is necessary for the pivot type provider.
The typing rules given are written using the following judgement:
%
\begin{equation*}
L_1; \Gamma \vdash e : \tau; L_2
\end{equation*}

@@

\noindent
The class $C$ may use classes defined in $L_1$, but also delayed classes from $L$. This models 
laziness as $L$ is a function that may never be evaluated. Since $L$ is potentially infinite, we 
cannot check class definitions upfront as in typical object calculi \cite{objects}. Instead, we check that 
that members are well typed as they appear in the source code, which matches the behaviour of F\# 
type providers. In general, this means that $L$ may contain classes with incorrectly typed 
members. We prove that this is not the case for the pivot type provider (Section~\ref{sec:pivot-prop}).

@@

The rules for primitive types and variables are standard. Input data (\emph{data}) is of type 
\ident{Query} and all the operations of relational algebra take a value of type \ident{Query} and 
produce a result of type \ident{Query}. An untyped \ident{Query} value can be converted into a 
series type (\emph{series}) of any type -- in the pivot type provider, the \ident{Query} type is 
always hidden and users can only work with type-safe series type. The bridging of typed and 
untyped parts of the language resembles the boundary between static and dynamic typing in 
gradually typed languages~\cite{gradual}.

@@

In an implementation, a type provider is an executable component that is called by the compiler
and the editor to provide information about types on demand. In our formalization, we follow the 
style used by Petricek et al. \cite{fsdata}, but we add laziness as discussed in Section~\ref{sec:foo-foo}.
We model the core operations (dropping columns, grouping and sorting) in Section~\ref{sec:pivot-core}
and then refine the model to include filtering based on values Section~\ref{sec:pivot-filter}.
We omit paging, as it does not affect the shape of data, but modelling it would require richer 
host language.

@@

\noindent
In prior work on type providers \cite{fsdata}, the actual input value is not available when writing the code
and so the schema is inferred from a representative sample value. In exploratory data analysis, the 
data set is often available at the time of writing the code and so $D$ can be the actual data set.
However, it could equally be a sample or representative subset of the data.

@@

\subparagraph{Spreadsheet-inspired live editor.}
The third editor extension that we developed for the pivot type provider aims to bridge the gap 
between code and user interface. This is done through a direct manipulation interface \cite{directman} 
inspired by spreadsheet applications. When exploring data in a spreadsheet, the user can always see the 
data they work with and the results of an action will be immediately visible. For queries written
using the pivot type provider, the results can also be calculated immediately using the sample 
data set provided when using the refined version of the type provider as discussed in 
Section~\ref{sec:pivot-filter}.

@@

Unlike the user interface for modifying lists, the live editor works specifically with the pivot 
type provider. It still relies on the simple structure provided by the fact that entire 
transformation can be written as a single chain of member accesses. In particular,
we identify individual transformations (\qident{group by}, \qident{sort by}, etc.) and 
generate different user interface for specifying parameters of each transformation. For sorting, as 
shown in Figure~\ref{fig:case-ed}, the user can add or remove sort keys. For grouping 
or paging, the user interface lets the user choose the grouping key and the number of 
elements to take, respectively.

@@

Although the technical focus of this paper is on the programming language theory behind the pivot type
provider (Section~\ref{sec:pivot}), the paper also outlines interesting human-computer interaction 
aspects (Section~\ref{sec:impl}). We discuss further related directions in this
section before~concluding.

\subsection{Related and further work}
\label{sec:further}

\subparagraph{Type providers.}
Type providers have been pioneered in F\# \cite{inforich}. They can be seen as a form of dependent 
typing \cite{idris-tp}; we take the opposite perspective and use type providers as a mechanism for 
implementing other type system features. Our focus on using type providers for describing 
computations is different from other type provider work \cite{fsdata,liteq,ageweb}, which focuses 
on mapping of external data into types.

\subparagraph{Fancy types.}
The pivot type provider makes data exploration safer as it does not allow construction of invalid
queries. Alternative approach would be to use fancy types \cite{fancytypes}. This can be done
using using row types and typestate or phantom types \cite{rowtypes,typestate,phanty} as sketched
in in Section~\ref{sec:columns-row}. The idea of using type providers in this way has also been
explored for session types \cite{sessionty,sessiontp} and it would be interesting to see whether
our approach can be applied in other areas such as web development \cite{urweb}.

\subparagraph{Human-computer intercation.}
We discussed how the pivot type provider simplifies the programming model (Section~\ref{sec:analysis}),
but it would be interesting to explore this aspect empirically through the perspective of HCI. 
The live editor shown in Section~\ref{sec:impl} offers a form of direct manipulation 
\cite{directman,directman2,dynamicq}. Unlike spreadsheets, we construct a transformation rather than
actually transforming data, which makes it more related to systems for query construction
\cite{spreadsheetalgebra,querydirect}. We also see code as equally important to the direct 
manipulation interface.

\subparagraph{Relational algebra.}
Our operational semantics used to model data transformations (Section~\ref{sec:foo}) was based on 
relational algebra \cite{relalg,dbsys}, although our focus was on aggregation, which has been 
added to the core algebra in a number of different ways \cite{sumtables,datacube,relalg-alpha,relalg-sparql}.
The pivot type provider does not provide operations for joining data sets, which is an interesting
problem for further work as it requires extensions to the type provider mechanism -- the join
operation is parameterized by two data sets that are being combined.

\subsection{Conclusions}
There is lots of data (big data), but using them is hard (post-truth).
We want to make it easier to explore data without sacrificing reproducibility 
and other good things we get from having source code.

In this paper, we focus on making the language as simple as possible. We unify many different
PL concepts using member access as the primary programming mechanism thanks to type providers.
This makes programming easier through auto-completed, but we also explored other tooling ranging
from simple interactive UIs to direct manipulation tools.

The pivot type provider provides a safe and easy to use layer over an underlying 
relational algebra. We formalize the host language and the type provider and we prove
that queries constructed using the type provider are correct.

TL;DR - better toools to democratize data stuff.

\bibliography{paper}

@@
